---
title: "Estimate survival times - Leuk data"
author: "Naga Pakalapati"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message=F,warning=F,echo=F,fig_height=10,fig_width=7,cache = F)
```

```{r}
#List of libraries used in the project
library(HSAUR3)
library(knitr)
library(ggplot2)
library(broom)
library(patchwork)
```

### Estimate survival times


The \textbf{leuk} data from package \textbf{MASS} shows the survival times from diagnosis of patients suffering from leukemia and the values of two explanatory variables, the white blood cell count (wbc) and the presence or absence of a morphological characteristic of the white blood cells (ag).


- We define a binary outcome variable according to whether or not patients lived for at least 24 weeks after diagnosis and call it \textit{surv24}. 
    
- Fit a logistic regression model to the data with \textit{surv24} as response.
    
- We shall construct some graphics useful in the interpretation of the final model.
    
- Fit a model with an interaction term between the two predictors and which model fits the data better?
    
Let's define a binary outcome as suggested and fit a logistic regression model.
    
```{r}
#load data
data("leuk", package = "MASS")
leuk_dat <- leuk

#adding surv24 column
leuk_dat$surv24 <- ifelse(leuk_dat$time>=24, 1,0)

#fit glm (note: wbc is logtransformed with base 10)
model1 <- surv24 ~ log10(wbc) + ag
glm_m1 <- glm(model1, data = leuk_dat, family = binomial())

summary(glm_m1)

```

From the above summary of the model, we can see that **wbc** is not siginificant and the
presence of morphologic characteristic of wbc is very significant.
    
Let's visualize how the probability of longer life expectancy is based on ag present or 
absent.

```{r}
#plot probabiliy of ag for each category
probability <- predict(glm_m1, leuk_dat, type = "response")

plot(probability~leuk_dat$ag, main="Probability of longer life expectancy (base plot)",
     xlab="presence or absence of a morphological characteristic of the white blood cells",
     col=c("red", "green"))

#plot in ggplot
ggplot() +
  geom_boxplot(aes(x=leuk_dat$ag, y=probability),color = c("red", "green")) +
  ggtitle("Probability of longer life expectancy (ggplot)") +
  xlab("presence or absence of a morphological characteristic of the white blood cells")
```

    
Lets' fit a second model with interation term.
    
```{r}
#fit glm with interaction term
model2 <- surv24 ~ log10(wbc) + ag + log10(wbc)*ag
glm_m2 <- glm(model2, data = leuk_dat, family = binomial())
summary(glm_m2)
```

    
We fitted two models, one with wbc and ag predictoers and another with these both and the
interaction term between these two. We find that there is no significant interation between
the terms. So the model with just the original predictors fits the data better.

Also, the higher count of wbc doesn't seem to significantly affect the longivity of survival.
But the presence or absence of a morphological characteristic of the white blood cells (ag)
seems to be the major factor. We can say that the presence of (ag) will increase the number of
weeks of survival.

---------------
